Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-14 paper/paper.tex
working on revisions for paper

Kiselyov \cite{kiselyov:sharing}  presents a method for implementing
eDSLs in finally tagless form that generate a directed acyclic graph (DAG) with
sharing. 
However, as we will explain in sections~\ref{limithashcons} and
\ref{limitexplicit}, for DSL functions that return multiple
outputs (e.g., tuples, lists, etc.), Kiselyov's method of detecting sharing may require computation exponential in the size of the program,
and his method of explicitly declaring sharing is inapplicable.

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} exhibits this problem.
 In our work,
this translated into the inability to process large library functions.

@@

We can then create different instances to implement different functionality.
For example, we can implement a pretty printer for our AST as
\begin{minted}{haskell}
newtype Pretty a = Pretty { unPretty :: String }

@@

@@

Kiselyov's method
 for detecting implicit sharing in finally tagless
style uses hash-consing \cite{kiselyov:sharing}.
Hash-consing is based on a bijection of nodes and a set of identifiers,
e.g., with interface 
\begin{minted}{haskell}
data BiMap a -- abstract
lookup_key :: Ord a => a -> BiMap a -> Maybe Int
lookup_val :: Int -> BiMap a -> a
insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
empty :: BiMap a
\end{minted}
An efficient implementation using hashing and linear probing is given by Thai
in his Master's thesis \cite{thai2021type}.

Nodes need to be uniquely identifiable, and shouldn't be a recursive data type,
such as
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int
\end{minted}
The representation for the finally tagless
instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
returns the current (top) \mintinline{haskell}{NodeID}:
\begin{minted}{haskell}
newtype DAG = DAG (BiMap Node) deriving Show

@@

When we wrap our state monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
\begin{minted}{haskell}
haskellSharing x y=
 let
   z = x + y
 in z + z
\end{minted}

>>>>

Kiselyov \cite{kiselyov:sharing} presents a method for implementing eDSLs in
finally tagless form that generate a directed acyclic graph (DAG) with sharing.
However, as we will explain in sections~\ref{limithashcons} and
\ref{limitexplicit}, for DSL functions that return multiple outputs (e.g.,
tuples, lists, etc.), Kiselyov's method of implicitly detecting sharing may
require computation exponential in the size of the program, and his method of
explicitly declaring sharing is inapplicable.

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
it's two outputs when writing it's implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

@@

We can then create different instances to implement different functionality.
For example, we can implement a pretty printer
\begin{minted}{haskell}
newtype Pretty a = Pretty { runPretty :: String }

@@

Or generate an AST
\begin{minted}{haskell}
newtype AST a = AST { genAST :: Exp }

instance Exp AST where
  add x y =  AST $ Add (genAST x) (genAST y)
  variable x = AST $ Variable x
  constant x = AST $ Constant x
\end{minted}

@@

The goal of detecting sharing is to generate a graph structure (like the AST in
the previous section) that share's common subexpressions. So we're essentially
going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example

\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int

newtype DAG = DAG (BiMap Node) deriving Show
\end{minted}

Kiselyov's method for detecting implicit sharing in finally tagless style uses
hash-consing \cite{kiselyov:sharing}. Hash-consing is based on a bijection of
nodes and a set of identifiers, e.g.,
\begin{minted}{haskell}
data BiMap a -- abstract
lookup_key :: Ord a => a -> BiMap a -> Maybe Int
lookup_val :: Int -> BiMap a -> a
insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
empty :: BiMap a
\end{minted}

The method can be performed using any data structure that provides the above
interface. An efficient implementation would use hashing and linear probing, as
is done by Thai in his Master's thesis \cite{thai2021type}.

The representation for the finally tagless
instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
returns the current (top) \mintinline{haskell}{NodeID}:
\begin{minted}{haskell}

@@

When we wrap our state monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
\begin{minted}{haskell}
haskellSharing x y =
 let
   z = x + y
 in z + z
\end{minted}
